---
title: Lab 1
---

## Types of problems

1. **Modeling with networks** Of the following cases, which ones can be treated with "regular statistical methods"? Justify your answers by identifying what is the unit of analysis in your "regression"/"test" and specify whether the units are independent or not.

a. Egocentric study 1: Analyze how the network structure affects health outcomes.

b. Egocentric study 2: Investigate what network structures are more prevalent in a population sample.

c. Egocentric study 3: Elucidate whether the prevalence of a given network structure is higher in one population than in another.

d. Country-level networks: Analyze whether neighboring countries tend to adopt international treaties at the same time.

e. Phylogenomics: Study a given phenotype in a population of organisms related by a phylogenetic tree. 



## Programming 

1. **Simulating convolutions**: Using what you have learned about statistical functions in R, simulate the convolution of two normal distributions, one with $(\mu, \sigma^2) = (-3, 1)$ and the other with $(\mu, \sigma^2) = (2, 2)$. Plot the histogram of the results. Draw 1,000 samples.

```{r}
#| label: simulations-convolutions
#| eval: false
#| echo: false
set.seed(1)
x <- rnorm(1000, mean = -3, sd = 1)
y <- rnorm(1000, mean = 2, sd = 2)
z <- x + y

hist(z)
```

2. **Bimodal distribution**: Using the previous two normal distributions, simulate a bimodal distribution where the probability of sampling from the first distribution is 0.3 and the probability of sampling from the second distribution is 0.7. Plot the histogram of the results. (**Hint:** use a combination of `runif()` and `ifelse()`).

```{r}
#| label: simulations-bimodal
#| eval: false
#| echo: false
z <- ifelse(runif(1000) < 0.3, x, y)
density(z) |> plot()
```

## Random graphs

1. Write a function to simulate Bernoulli graph, but instead of evaluating each $(i,j)$ tie individually like in @er-code, use a binomial random number to identify how many edges the graph will have, and randomly pick which will those be from the set $V\times V$

2. Using a Generalized-Linear-Model [GLM], estimate the density parameter of the previous graph.